---
title: "Things to consider adding"
author: "Gavin Masterson"
date: "08/11/2020"
output: html_document
---

#### Does this matter? 

(It is important to highlight that the number shown above indicates how many times non-singletons occurred at a single trap array, even if they occurred on the same date.)
To see how many unique days I captured more than a singleton of each of these species, we need to group by date as shown below.

```{r}
sitephase_summary %>% 
  filter(count_per_day > 1) %>% 
  group_by(species, date) %>% 
  summarise(count_per_day = n(), .groups = "drop_last") %>% 
  summarise(count_of_non_singleton_days = n(), .groups = "drop") %>% 
  arrange(desc(count_of_non_singleton_days))
```

This would only be relevant for models which group all the detection data for species by day and ignore site-specific detections.


```{r}
(sitephaseday_summary <- data %>%
  count(trap_site, 
        species,
        phase, 
        date,
        name = "count_per_day") %>%
  filter(count_per_day > 1) %>% 
  group_by(species) %>% 
  summarise(count_of_non_singleton_days = n(),
         max_captures = max(count_per_day)) %>% 
  arrange(desc(count_of_non_singleton_days)))
```
