---
title: "Having some fun exploring"
author: "Gavin Masterson"
date: "2020-10-12"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## **Introduction**

The raw data from my survey have been captured, cleaned and are now ready to be viewed for different insights.
As the survey was quite complex from a methodological and practical perspective, I will now produce summaries and visualisations of the relevant survey components and variables.
These will help me in determining any issues that may influence my desired analysis path.
Keeping with best practice, we first load the required `R` packages and the `data` we will use for all of exploratory analysis.

```{r setup, message = FALSE, echo = TRUE}
library(tidyverse)
library(lubridate)
library(here)

data <- read_csv(here("/data/clean_data.csv")) %>% 
           mutate(date = as_date(date),
                  arm = as_factor(arm),
                  block = as_factor(block),
                  phase = factor(phase, levels = c("pre-treatment", "post-treatment")))
```

#### **The total number of species detected during the survey**

Suikerbosrand Nature Reserve has been surveyed for reptiles several times since its proclamation in 1973 e.g. Jacobsen *et al.* (1986) and Masterson *et al.* (2008).
From these surveys, we know that there are roughly 40 species of reptile that have ever been recorded in Suikerbosrand Nature Reserve during its history.
Determining the total number of species detected during this survey is an important first step in gauging the general success of the survey. 

```{r total species}
data %>% 
  drop_na(species) %>% 
  summarise(total_species = n_distinct(species))
```

Twenty species is just half of the estimated 40 reptile species that might be present in the reserve, so how well did we really do?
While we might have hoped to capture a greater number of the species, there are two reasons that we need not be too concerned with only having caught 21 species. 
Firstly, the survey was not aiming to catch all the species in the reserve.
This survey aimed to understand the effects of disturbance on the local assemblages in one type of habitat - open grassland. 
Secondly, there is no reason to expect that all 40 species that have ever been recorded in Suikerbosrand Nature Reserve are still present in the reserve. 
The most intense survey of the reserve was conducted by Niels Jacobsen in the 1980's and the surrounding area, now called Gauteng Province, has undergone rapid intensification of land use in the past 30+ years.
Therefore, the positive view of this survey is that we can confidently say that at least 21 reptile species still occur in Suikerbosrand Nature Reserve.

#### **The total number of captures per day when traps were checked**

```{r total captures per day}
zero_dates <- data[which(is.na(data$species) == TRUE), 1:2] %>%
  mutate(total_captures = as.integer(0))

data %>%
  drop_na(species) %>% 
  group_by(date, phase, trap_type) %>% 
  summarise(total_captures = n(),
            .groups = "drop") %>%
  full_join(zero_dates) %>%
  arrange(date) %>% 
  ggplot(., mapping = aes(x = date,
                          y = total_captures)) +
    geom_col(aes(fill = trap_type)) +
    xlab("Date")  +
    guides(x = guide_axis(angle = 90)) +
    ylab("Total captures") +
    labs(fill = "Trap type") +
    scale_fill_discrete(na.translate=FALSE)
```

#### **The total number of detections per species**

```{r captures per species}
data %>%
  drop_na(species) %>% 
  group_by(species, phase) %>% 
  summarise(total_captures = n(),
            .groups = "drop_last") %>% 
  arrange(species) %>% 
  ggplot(., mapping = aes(x = species,
                          y = total_captures)) +
    geom_col(aes(fill = phase),
             position = position_dodge()) +
    xlab("Species") +
    guides(x = guide_axis(angle = 90)) +
    ylab("Total captures") +
    labs(fill = "Phase")
```

#### **The total number of species detected at each site**

```{r species per site}
sitephase_summary <- data %>%
  group_by(trap_site, species, phase) %>%
  summarise(count_per_trap_site = n())
```

#### **Species captures per site per day**

One of the considerations when using hierarchical models is whether to use detection or abundance data to model species occupancy.
Reptiles do not have to be active every day like mammals, which means that there are unlikely to be multiple encounters with trap arrays on a daily basis. 
As a result, I expected to catch just a single individual per trap array per day of checking.

```{r}
# create pivot table of species per day, per trap array and phase
sitephase_summary <- data %>%
  group_by(trap_site, species, phase, date) %>%
  summarise(count_per_day = n(), .groups = "drop")

sitephase_summary %>% 
  filter(count_per_day > 1) %>% 
  group_by(species) %>% 
  summarise(count_of_non_singleton_days = n(), .groups = "drop") %>% 
  arrange(desc(count_of_non_singleton_days))
```

This summary shows me the seven species which did not match my expectation, in descending order of non-singleton days.
The first five are the common lizard species in Suikerbosrand Nature Reserve, and the last two are snake species.
One other question we might ask is whether or not the number of non-singleton days was different between the pre-treatment and post-treatment phases of the project. 

```{r non-singleton days per phase}
sitephase_summary %>% 
  filter(count_per_day > 1) %>% 
  group_by(species,phase) %>% 
  summarise(count_of_non_singleton_days = n(), .groups = "drop") %>% 
  arrange(desc(count_of_non_singleton_days))
```

Incredibly, we see a near-perfect, 50-50 split of non-singleton days between the two phases for four of the species. 
The  three species for which the split is not even are *P. wahlbergii* (pre 0:5 post), *A. a. distanti* (pre 2:0 post) and *H. haemachatus* (pre 0:1 post).
Of these three species, only *A. a. distanti* is deserving of further consideration.
Knowing that the species was infrequently captured during the survey - with zero captures in the post-treatment phase of the project - it is important to consider how much data could be lost if we reduce these non-singleton days to singletons for use in an 'any detection per day' formulation of our hierarchical model.

```{r}
sitephase_summary %>% 
  filter(species == "Agama aculeata distanti") %>% 
  ggplot(., mapping = aes(x = date,
                          y = count_per_day)) +
    geom_col(aes(fill = trap_site),
             position = "dodge") +
    xlab("Date")  +
    guides(x = guide_axis(angle = 90)) +
    ylab("Total captures") +
    labs(fill = "Trap site")
    

```

This shows that I captured three individuals at `D1` on both of the non-singleton days, which were only two days apart in October 2005.
This means that six detections of *A. a. distanti* would be reduced to two detections using a 'detection per site per day' approach, and reduce the total number of detections from 25 to 21 - a 16 percent reduction.











#### Does this matter? 

(It is important to highlight that the number shown above indicates how many times non-singletons occurred at a single trap array, even if they occurred on the same date.)
To see how many unique days I captured more than a singleton of each of these species, we need to group by date as shown below.

```{r}
sitephase_summary %>% 
  filter(count_per_day > 1) %>% 
  group_by(species, date) %>% 
  summarise(count_per_day = n(), .groups = "drop_last") %>% 
  summarise(count_of_non_singleton_days = n(), .groups = "drop") %>% 
  arrange(desc(count_of_non_singleton_days))
```

This would only be relevant for models which group all the detection data for species by day and ignore site-specific detections.

