---
title: "Having some fun exploring"
author: "Gavin Masterson"
date: "2020-10-12"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup}
library(tidyverse)
```

## Introduction

The raw data from my survey have been captured, cleaned and are now ready to be viewed for different insights.
As the survey was quite complex from a methodological and practical perspective, I will now produce summaries and visualisations of the relevant survey components and variables.
These will help me in determining any issues that may influence my desired analysis path. 
(@) The total number of species detected during the survey  
Suikerbosrand Nature Reserve has been surveyed for reptiles several times since its proclamation in 1973 e.g. Jacobsen *et al.* (1986) and Masterson *et al.* (2008).
Determining the total number of species detected during this survey is an important first step in gauging the general success of the survey.  
```{r total species, include = TRUE}
data %>% 
  summarise(total_species = n_distinct(species))
```
(@) The total number of species detected at each site  

```{r species per site}
sitephase_summary <- data %>%
                      group_by(trap_site, species, phase) %>%
                      summarise(count_per_trap_site = n())
```
